---
title: "ROHR_07_Gut_stats_Beta_div"
author: "LLL"
date: "2022-11-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Project summary

- Fish were collected from 2 sites in the Tropical Eastern Pacific: Las Perlas & Coiba. 
- Sampling during both upwelling & non-upwelling seasons (collections done 2021-22). 
- 12 target species, 10 of which are represented in this dataset. 
- Gut = 8 per species.
- Control swabs (swabs exposed to air in field) collected at each site and season
- Swabs were stored in liquid nitrogen for transport from field to lab, then at -80ºC
- DNA was extracted using ZymoBIOMICS 96 well DNA kit - 4 plates total (per sample type = skin & gut)
- DNA was amplified following a modified version of the Earth Microbiome Project 16S protocol using phased primers (1 set per plate)
- Negative PCR controls and extraction controls included in each plate
- Index PCR performed to attach unique barcodes, then all four plates (384 samples) pooled for sequencing on the Illumina MiSeq sequencing platform of the Smithsonian Tropical Research Institute's (STRI) Naos facilities in Panama (courtesy of Marta Vargas)
- Sequenced libraries were demultiplexed at STRI by Marta Vargas using the MiSeq Reporter Software
- These libraries were sent to me (Laura Lardinois) and trimmed with Cutadapt to remove primers before being read into R to start the DADA2 pipeline (following tutorial v 1.16) 
- Silva version 138.1 (mar 10, 2021 update) used for taxonomic assignments
- See "RRR Fish 16S Sequences Summer 2022" document for pipeline and processing up to the creation of the initial phyloseq object

- This code was adapted from various source files including Matt Leray's BCS work, the Bocasbiome project, and several tutorial pages



Loading our phyloseq object - filtered & trimmed (not rarefied) with phylogenetic tree - from "ROHR_07_Gut_stats_alpha.Rmd"
```{r read phyloseq object - Gut only, echo=FALSE}
ROHR_07_obj_phylo_root <- readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_Gut_with_phylogenetic_tree_rooted.rds")
```


```{r load required packages, echo=FALSE}
library(dada2)
library(ggplot2)
library(ff)
library(phyloseq)
library(gridExtra)
library(dplyr)
library(decontam)
library(ape)
library(reltools)
library(ampvis2)
library(edgeR)
library(PathoStat)
library(ape)
library(phangorn)
library(seqinr)
library(paco)
library(MCMCglmm)
library(phytools)
library(vegan)
library(reshape2)
library(pairwiseAdonis) #for pairwise comparisons following PERMANOVA (adonis2 function in vegan)
library(GUniFrac) #to perform Generalized UniFrac Distance calculation
library(hillR) #hill numbers package - calculate taxonomic, functional, and phylogenetic diversity via Hill's #s
library(forcats) #to reorder samples for visualization (e.g., by trophic group)
```


############Beta Diversity Metrics

To calculate Beta diversity, we need to normalize for sample depth - many ways to do this, including transforming counts into relative abundances (may not be the best way - see https://astrobiomike.github.io/amplicon/dada2_workflow_ex#assigning-taxonomy)

Beta diversity workflow from Bocasbiome project: https://bocasbiome.github.io/wf5.html

```{r counts to relative abundances}
####### Transform counts to relative abundance per sample
ROHR_07_obj_clean_prop = transform_sample_counts(ROHR_07_obj_phylo_root, function(x) x/sum(x))
```

PERMANOVA (Permutational Multivariate Analysis of Variance)

source: https://archetypalecology.wordpress.com/2018/02/21/permutational-multivariate-analysis-of-variance-permanova-in-r-preliminary/

- non-parametric, multivariate statistical test
- compares groups of objects
H0 (general) = centroids & dispersion of groups are equivalent for all groups
H0 (specific to this case) = There are no differences in the relative abundances of microbial taxa in the fish gut microbiome across all sites, species and seasons in the Tropical Eastern Pacific. 

-test statistic = pseudo-F ratio
"It compares the total sum of squared dissimilarities (or ranked dissimilarities) among objects belonging to different groups to that of objects belonging to the same group. Larger F-ratios indicate more pronounced group separation, however, the significance of this ratio is usually of more interest than its magnitude." 

Assumptions:
- objects in dataset are exchangeable under the null hypothesis
- exchangeable objects (sites, samples, observations) are independent ***** (need to account for this assumption NOT being met in test setup)
- exchangeable objects have similar multivariate dispersion (betadisper --> multivariate analogue to Levene's test for homogeneity of variances) (see code chunk below)

```{r Permanova fish gut microbiome full dataset - Bray-Curtis distances}
#before running PERMANOVA, make sure DESeq2 is not loaded, as it causes an issue
#detach(package:DESeq2, unload=TRUE)

####running the PERMANOVA
metadata <- as(sample_data(ROHR_07_obj_clean_prop), "data.frame")

######### by species ######

###species (fish host species) (unconstrained permutations)
adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ species,
        data = metadata)

#adonis2(formula = distance(ROHR_07_obj_clean_prop, method = "bray") ~ species, data = metadata)
#          Df SumOfSqs      R2      F Pr(>F)    
#species    9   49.327 0.35184 18.215  0.001 ***
#Residual 302   90.869 0.64816                  
#Total    311  140.196 1.00000

# species = highly significant (p < 0.001), R2 = 0.36184


####constraining permutations w/ strata to only run within a region (control for effect of region --> see if relative abundances are different w/in sites between species)
with( metadata, adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ species,
        data = metadata, strata = region)) ##no effect of using strata --> is this actually working??? may need to use "permute" package to do this

#adonis2(formula = distance(ROHR_07_obj_clean_prop, method = "bray") ~ species, data = metadata, strata = region)
#          Df SumOfSqs      R2      F Pr(>F)    
#species    9   49.327 0.35184 18.215  0.001 ***
#Residual 302   90.869 0.64816                  
#Total    311  140.196 1.00000               

######## by region #######
            
###region (collection region - Coiba or Las Perlas)
adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ region,
        data = metadata)

#adonis2(formula = distance(ROHR_07_obj_clean_prop, method = "bray") ~ region, data = metadata)
#          Df SumOfSqs      R2      F Pr(>F)
#region     1    0.604 0.00431 1.3408  0.124
#Residual 310  139.592 0.99569              
#Total    311  140.196 1.00000 
                  

####by site nested within region (collection region - Coiba or Las Perlas) --> accounting for nestedness (sampling sites found w/in regions)

##confirm correct order for nestedness --> region/site or site/region?

with(metadata, adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ region/site,
        data = metadata, by = "margin"))

#adonis2(formula = distance(ROHR_07_obj_clean_prop, method = "bray") ~ region/site, data = metadata, by = "margin")
#             Df SumOfSqs      R2      F Pr(>F)    
#region:site   9    6.693 0.04774 1.6843  0.001 ***
#Residual    301  132.899 0.94795                  
#Total       311  140.196 1.00000  

##significant interaction between region and site

####### by season ######### 
adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ season,
        data = metadata, strata = metadata$region) #need to constrain permutations w/in region? 

#adonis2(formula = distance(ROHR_07_obj_clean_prop, method = "bray") ~ season, data = metadata, strata = metadata$region)
#          Df SumOfSqs      R2      F Pr(>F)
#season     1    0.413 0.00294 0.9151  0.566
#Residual 310  139.783 0.99706              
#Total    311  140.196 1.00000  

#season alone = NSD

#### by date nested within season
with(metadata, adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ season/date,
        data = metadata, strata = season)) 

#adonis2(formula = distance(ROHR_07_obj_clean_prop, method = "bray") ~ season/date, data = metadata, strata = season)
#             Df SumOfSqs      R2      F Pr(>F)    
#season        1    0.413 0.00294 0.9566  0.001 ***
#season:date  24   16.421 0.11713 1.5863  0.001 ***
#Residual    286  123.362 0.87993                  
#Total       311  140.196 1.00000 

####season*region/site
with(metadata, adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ region/site*season,
        data = metadata, strata = region, by = "margin"))

#adonis2(formula = distance(ROHR_07_obj_clean_prop, method = "bray") ~ region/site * season, data = metadata, by = "margin", strata = region)
#                    Df SumOfSqs      R2      F Pr(>F)    
#region:site:season   3    2.605 0.01858 1.9856  0.001 ***
#Residual           296  129.442 0.92329                  
#Total              311  140.196 1.00000 

#significant interaction between region, site, and season  

####season*region/site + species
with(metadata, adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ region/site*season + species,
        data = metadata, strata = region, by = "margin"))

#adonis2(formula = distance(ROHR_07_obj_clean_prop, method = "bray") ~ region/site * season + species, data = metadata, by = "margin", strata = region)
#                    Df SumOfSqs      R2       F Pr(>F)    
#species              9   44.453 0.31707 16.6791  0.001 ***
#region:site:season   3    1.100 0.00784  1.2379  0.091 .  
#Residual           287   84.989 0.60622                   
#Total              311  140.196 1.00000   

#testing for interactions b/w species, region (w/ site nested w/in region), and season
with(metadata, adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ species*region/site*season,
        data = metadata, strata = region))

#adonis2(formula = distance(ROHR_07_obj_clean_prop, method = "bray") ~ species * region/site * season, data = metadata, strata = region)
#                            Df SumOfSqs      R2       F Pr(>F)    
#species                      9   49.327 0.35184 19.7543  0.001 ***
#region                       1    0.638 0.00455  2.2984  0.001 ***
#season                       1    0.414 0.00295  1.4909  0.058 .  
#species:region               9    4.133 0.02948  1.6551  0.001 ***
#species:season               9    3.582 0.02555  1.4347  0.001 ***
#region:season                1    0.498 0.00355  1.7963  0.010 ** 
#species:region:site         38   12.186 0.08692  1.1558  0.002 ** 
#species:region:season        8    2.634 0.01879  1.1868  0.033 *  
#species:region:site:season   9    4.082 0.02912  1.6348  0.001 ***
#Residual                   226   62.703 0.44725                   
#Total                      311  140.196 1.00000 


####season/date*region/site + species
with(metadata, adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ region/site*season/date*species,
        data = metadata, strata = region))

#adonis2(formula = distance(ROHR_07_obj_clean_prop, method = "bray") ~ region/site * season/date * species, data = metadata, strata = region)                                 
#                                 Df SumOfSqs      R2       F Pr(>F)    
#region                            1    0.604 0.00431  2.1599  0.001 ***
#season                            1    0.413 0.00294  1.4759  0.065 .  
#species                           9   49.362 0.35209 19.6212  0.001 ***
#region:site                       9    3.286 0.02344  1.3063  0.002 ** 
#region:season                     1    0.443 0.00316  1.5832  0.030 *  
#region:species                    9    3.921 0.02797  1.5587  0.001 ***
#season:species                    9    3.472 0.02476  1.3800  0.001 ***
#region:site:season                3    1.142 0.00815  1.3622  0.022 *  
#region:site:species              29    9.410 0.06712  1.1609  0.002 ** 
#region:season:species             8    3.084 0.02200  1.3793  0.001 ***
#region:site:season:date          19    5.415 0.03862  1.0195  0.325    
#region:site:season:species        6    2.240 0.01598  1.3358  0.003 ** 
#region:site:season:date:species  31    8.208 0.05854  0.9472  0.865    
#Residual                        176   49.197 0.35091                   
#Total                           311  140.196 1.00000  

##how to deal with all these interactions in the post-hoc tests?                
```

```{r betadisper - check dispersion}
ROHR_07_bray <- phyloseq::distance(ROHR_07_obj_clean_prop, method = "bray")

#homogeneity of dispersion test
dispersion <- betadisper(ROHR_07_bray, metadata$species)
permutest(dispersion) #see that P <0.001, meaning that betadisper results ARE significant and we can reject our null hypothesis that the species have the same centroid --> means that the PERMANOVA results may be influenced by differences in group dispersion, rather than truly reporting significant differences between group centroids

plot(dispersion, hull=FALSE, ellipse=TRUE) ##sd ellipse
#this shows us that A. concolor is a big outlier in terms of dispersion, with much greater dispersion than any of the other species
```

For certain b-diversity analyses such as Jaccard, we will begin by rarefying (normalizing) our data to at least 1,000 reads (this will exclude 9 more samples). This cutoff can be modified to exclude fewer/more samples with low sequencing depth

```{r normalizing data}
#check current reads/sample
reads_per_sample <- sample_sums(ROHR_07_obj_phylo_root)

#how many samples with fewer than 1000 reads?
print(length(reads_per_sample[reads_per_sample < 1000])) #2 samples will be excluded under this criterion

set.seed(9242)  # This will help in reproducing the filtering and normalisation. 

ROHR07.rar.P <- rarefy_even_depth(ROHR_07_obj_phylo_root, sample.size = 1000)

#1143OTUs were removed because they are no longer present in any sample after random subsampling

```

```{r save rarefied data as new phyloseq object}
# Save rarefied data
saveRDS(ROHR07.rar.P, "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_rarefied_1000_P.rds")

#read rarefied data
ROHR07.rar.P <- readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_rarefied_1000_P.rds")

#check how much data post-rarefaction
print(ROHR07.rar.P) # 2075 taxa, 310 samples
```
```{r Permanova - Bray Curtis & Jaccard distances comparison}

####running the PERMANOVA
metadata <- as(sample_data(ROHR_07_obj_clean_prop), "data.frame")
metadataJac <- as(sample_data(ROHR07.rar.P), "data.frame")

######### by species ######

####Bray-Curtis
###species (fish host species) (unconstrained permutations)
adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ species,
        data = metadata)


####Jaccard

####can Jaccard be run with relative abundance data? or do we need to use raw count or rarefied count data?

###species (fish host species) (unconstrained permutations)
adonis2(distance(ROHR07.rar.P, method="jaccard") ~ species,
        data = metadataJac)


######## by region #######

####Bray-Curtis            
###region (collection region - Coiba or Las Perlas)
adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ region,
        data = metadata)


####Jaccard
###region (collection region - Coiba or Las Perlas)
adonis2(distance(ROHR07.rar.P, method="jaccard") ~ region,
        data = metadataJac)                  



####### by season ######### 

####Bray-Curtis 
adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ season,
        data = metadata)


####Jaccard 
adonis2(distance(ROHR07.rar.P, method="jaccard") ~ season,
        data = metadataJac)



####Bray-Curtis
####species + season*region/site
with(metadata, adonis2(distance(ROHR_07_obj_clean_prop, method="bray") ~ species + region/site*season,
        data = metadata))


####Jaccard
####species + season*region/site
with(metadata, adonis2(distance(ROHR07.rar.P, method="jaccard") ~ species + region/site*season,
        data = metadataJac))  

     
```



#########Looking at suite of Beta diversity metrics##########

- Bray-Curtis
- Jaccard
- UniFrac
- Generalized UniFrac (GUniFrac)
- Weighted UniFrac

```{r Bray Curtis distances full dataset}

####dissimilarity matrix####
set.seed(1911)
type.bray <- phyloseq::distance(ROHR_07_obj_clean_prop, method = "bray")
sampledf <- data.frame(sample_data(ROHR_07_obj_clean_prop))

####beta dispersion####

###by species
beta.bray.species <- betadisper(type.bray, sampledf$species,
                         type = "centroid", bias.adjust = TRUE)
permutest(beta.bray.species, pairwise = TRUE, permutations = 10000) # P <0.001 - variance is not homogeneous

plot(beta.bray.species, hull=FALSE, ellipse=TRUE)

###by region
beta.bray.region <- betadisper(type.bray, sampledf$region,
                         type = "centroid", bias.adjust = TRUE)
permutest(beta.bray.region, pairwise = TRUE, permutations = 10000) # P = 0.24 - variance is homogeneous

plot(beta.bray.region, hull=FALSE, ellipse=TRUE)

###by season
beta.bray.season <- betadisper(type.bray, sampledf$season,
                         type = "centroid", bias.adjust = TRUE) # P = 0.46 - variance is homogeneous
permutest(beta.bray.season, pairwise = TRUE, permutations = 10000)

plot(beta.bray.season, hull=FALSE, ellipse=TRUE)

####see code chunks above for PERMANOVAs####

####pairwise comparisons#### 

###by species
set.seed(1911)
pairwise.adonis(type.bray, factors = sampledf$species) # significant differences in relative abundances of taxa for certain pairwise comparisons (e.g., A. concolor:C. colonus, A. concolor:C. panamensis, M. dorsalis & A. concolor/A. troschelii/C. humeralis/E. labriformis/J. nigrirostris) but not others (e.g., C. colonus:C. panamensis, P. laticlavius:A. xanthopterus). In general, it seems that trophic group and/or host family may be correlated with similar relative abundances of taxa (with M. dorsalis as a notable exception to this). Among the interesting results, the very high P value between C. colonus (planktivore) and its carnivorous relative, C. panamensis, may hint at the importance of shared evolutionary history over feeding strategy, at least for these hosts (& this specific microbial diversity metric)

###by region
set.seed(1911)
pairwise.adonis(type.bray, factors = sampledf$region) #NSD b/w regions (p = 0.24)

###by season
set.seed(1911)
pairwise.adonis(type.bray, factors = sampledf$season) #NSD b/w seasons (p = 0.46)
```

```{r Jaccard dissimilarity full dataset}

####dissimilarity matrix####

set.seed(1911)
type.jaccard <- phyloseq::distance(ROHR07.rar.P, method = "jaccard", binary = T)
sampledf <- data.frame(sample_data(ROHR07.rar.P))

####beta dispersion####

###by species
beta.jaccard.species <- betadisper(type.jaccard, sampledf$species,
                         type = "centroid", bias.adjust = TRUE)
permutest(beta.jaccard.species, pairwise = TRUE, permutations = 10000) # P < 0.001 - variance is not homogeneous across species

plot(beta.jaccard.species, hull=FALSE, ellipse=TRUE) #visually, clear differences in dispersion, with P. laticlavius & E. labriformis standing out

###by region
beta.jaccard.region <- betadisper(type.jaccard, sampledf$region,
                         type = "centroid", bias.adjust = TRUE)
permutest(beta.jaccard.region, pairwise = TRUE, permutations = 10000) # P = 0.927 

plot(beta.jaccard.region, hull=FALSE, ellipse=TRUE)

###by season
beta.jaccard.season <- betadisper(type.jaccard, sampledf$season,
                         type = "centroid", bias.adjust = TRUE) # P = 0.766
permutest(beta.jaccard.season, pairwise = TRUE, permutations = 10000)

plot(beta.jaccard.season, hull=FALSE, ellipse=TRUE)

####PERMANOVAs####
##currently, have not included strata to constrain permutations

###by species
adonis.jaccard.species <- adonis2(type.jaccard ~ species, data = sampledf,
                          permutations = 10000)
adonis.jaccard.species #R2 = 0.21471, P < 0.001

###by region
adonis.jaccard.region <- adonis2(type.jaccard ~ region, data = sampledf,
                          permutations = 10000)
adonis.jaccard.region #R2 = 0.0055, P = 0.003

###by site nested within region
adonis.jaccard.sitereg <- adonis2(type.jaccard ~ site/region, data = sampledf,
                          permutations = 10000) 
adonis.jaccard.sitereg 

###by season
adonis.jaccard.season <- adonis2(type.jaccard ~ season, data = sampledf,
                          permutations = 10000)
adonis.jaccard.season

####pairwise comparisons#### 

###by species

set.seed(1911)
pairwise.adonis(type.jaccard, factors = sampledf$species) 

# significant differences in relative abundances of taxa for certain pairwise comparisons, but fewer than for Bray-Curtis.

###by region
set.seed(1911)
pairwise.adonis(type.jaccard, factors = sampledf$region) #NSD b/w regions (p = 0.34)

###by season
set.seed(1911)
pairwise.adonis(type.jaccard, factors = sampledf$season) #NSD b/w seasons (p = 0.54)
```


```{r UniFrac distances full dataset}

####dissimilarity matrix####

set.seed(1911)
type.unifrac <- phyloseq::distance(ROHR07.rar.P, method = "unifrac", weighted = F)
sampledf <- data.frame(sample_data(ROHR07.rar.P))

####beta dispersion####

###by species
beta.unifrac.species <- betadisper(type.unifrac, sampledf$species,
                         type = "centroid", bias.adjust = TRUE)
permutest(beta.unifrac.species, pairwise = TRUE, permutations = 10000) # P =

plot(beta.unifrac.species, hull=FALSE, ellipse=TRUE)


###by region
beta.unifrac.region <- betadisper(type.unifrac, sampledf$region,
                         type = "centroid", bias.adjust = TRUE)
permutest(beta.unifrac.region, pairwise = TRUE, permutations = 10000) # P =

plot(beta.unifrac.region, hull=FALSE, ellipse=TRUE)

###by season
beta.unifrac.season <- betadisper(type.unifrac, sampledf$season,
                         type = "centroid", bias.adjust = TRUE) # P =
permutest(beta.unifrac.season, pairwise = TRUE, permutations = 10000)

plot(beta.unifrac.season, hull=FALSE, ellipse=TRUE)

####PERMANOVAs####

###by species

set.seed(1911)

adonis.unifrac.species <- adonis2(type.unifrac ~ species, data = sampledf,
                          permutations = 10000)
adonis.unifrac.species

###by region
adonis.unifrac.region <- adonis2(type.unifrac ~ region, data = sampledf,
                          permutations = 10000)
adonis.unifrac.region

###by site nested within region
adonis.unifrac.sitereg <- adonis2(type.unifrac ~ site/region, data = sampledf,
                          permutations = 10000)
adonis.unifrac.sitereg

###by season
adonis.unifrac.season <- adonis2(type.unifrac ~ season, data = sampledf,
                          permutations = 10000)
adonis.unifrac.season

####pairwise comparisons#### 

###by species

set.seed(1911)
pairwise.adonis(type.unifrac, factors = sampledf$species) #significant differences (p-adj) b/w almost all host species (except A. conc; A. xanth, A. conc; A. trosch, A. xanth: A. trosh, J. nigri; A. trosh)

###by region
set.seed(1911)
pairwise.adonis(type.unifrac, factors = sampledf$region) # p-adj = 0.003

###by season
set.seed(1911)
pairwise.adonis(type.unifrac, factors = sampledf$season) # p-adj = 0.002
```

```{r GUniFrac distances full dataset}

####dissimilarity matrix####

#whole dataset
asv.tab<-otu_table(ROHR07.rar.P)
tree.fish<-phy_tree(ROHR07.rar.P)
fish.sample<-sample_data(ROHR07.rar.P)

groups.region.whole <- fish.sample$region
groups.season.whole <- fish.sample$season
species.whole <- fish.sample$species

unifracs <- GUniFrac(asv.tab, tree.fish,
                     alpha=c(0, 0.5, 1))$unifracs
d5.all <- unifracs[, , "d_0.5"]
type.gunifrac <- as.dist(d5.all)


####beta dispersion####

###by species
set.seed(1911)
beta.gunifrac.species <- betadisper(as.dist(d5.all), species.whole,
                             type = "centroid", bias.adjust = TRUE)
permutest(beta.gunifrac.species, pairwise = TRUE, permutations = 10000)

plot(beta.gunifrac.species, hull=FALSE, ellipse=TRUE)


###by region
set.seed(1911)
beta.gunifrac.region <- betadisper(as.dist(d5.all), groups.region.whole,
                             type = "centroid", bias.adjust = TRUE)
permutest(beta.gunifrac.region, pairwise = TRUE, permutations = 10000)

plot(beta.gunifrac.region, hull=FALSE, ellipse=TRUE)


###by season
set.seed(1911)
beta.gunifrac.season <- betadisper(as.dist(d5.all), groups.season.whole,
                             type = "centroid", bias.adjust = TRUE)
permutest(beta.gunifrac.season, pairwise = TRUE, permutations = 10000)

plot(beta.gunifrac.season, hull=FALSE, ellipse=TRUE)


####PERMANOVAs####

sampledf <- data.frame(sample_data(ROHR07.rar.P))

###by species

adonis.gunifrac.species <- adonis(type.gunifrac ~ species, data = sampledf,
                           permutations = 10000)
adonis.gunifrac.species

###by region
adonis.gunifrac.region <- adonis(type.gunifrac ~ region, data = sampledf,
                          permutations = 10000)
adonis.gunifrac.region

###by site nested within region
adonis.gunifrac.sitereg <- adonis(type.gunifrac ~ site/region, data = sampledf,
                          permutations = 10000)
adonis.gunifrac.sitereg

###by season
adonis.gunifrac.season <- adonis(type.gunifrac ~ season, data = sampledf,
                          permutations = 10000)
adonis.gunifrac.season


####pairwise comparisons#### 

###by species

set.seed(1911)
pairwise.adonis(type.gunifrac, factors = sampledf$species, p.adjust.m = "bonferroni") 


###by region
set.seed(1911)
pairwise.adonis(type.gunifrac, factors = sampledf$region, p.adjust.m = "bonferroni") 
###by season
set.seed(1911)
pairwise.adonis(type.gunifrac, factors = sampledf$season, p.adjust.m = "bonferroni") 
```

```{r WUniFrac distances full dataset}

####dissimilarity matrix####

set.seed(1911)
type.wunifrac <- phyloseq::distance(ROHR07.rar.P, method = "wunifrac")
sampledf <- data.frame(sample_data(ROHR07.rar.P))

####beta dispersion####

###by species
set.seed(1911)
beta.wunifrac.species <- betadisper(type.wunifrac, sampledf$species,
                             type = "centroid", bias.adjust = TRUE)
permutest(beta.wunifrac.species, pairwise = TRUE, permutations = 10000)

plot(beta.wunifrac.species, hull=FALSE, ellipse=TRUE)


###by region
set.seed(1911)
beta.wunifrac.region <- betadisper(type.wunifrac, sampledf$region,
                             type = "centroid", bias.adjust = TRUE)
permutest(beta.wunifrac.region, pairwise = TRUE, permutations = 10000)

plot(beta.wunifrac.region, hull=FALSE, ellipse=TRUE)


###by season
set.seed(1911)
beta.wunifrac.season <- betadisper(type.wunifrac, sampledf$season,
                             type = "centroid", bias.adjust = TRUE)
permutest(beta.wunifrac.season, pairwise = TRUE, permutations = 10000)

plot(beta.wunifrac.season, hull=FALSE, ellipse=TRUE)


####PERMANOVAs####

sampledf <- data.frame(sample_data(ROHR_07_obj_clean_prop))

###by species

set.seed(1911)
adonis.wunifrac.species <- adonis(type.wunifrac ~ species, data = sampledf,
                           permutations = 10000)
adonis.wunifrac.species


###by region
adonis.wunifrac.region <- adonis(type.wunifrac ~ region, data = sampledf,
                          permutations = 10000)
adonis.wunifrac.region

###by site nested within region
adonis.wunifrac.sitereg <- adonis(type.wunifrac ~ site/region, data = sampledf,
                          permutations = 10000)
adonis.wunifrac.sitereg

###by season
adonis.wunifrac.season <- adonis(type.wunifrac ~ season, data = sampledf,
                          permutations = 10000)
adonis.wunifrac.season


####pairwise comparisons#### 

###by species
set.seed(1911)
pairwise.adonis(type.wunifrac, factors = sampledf$species) 

###by region
set.seed(1911)
pairwise.adonis(type.wunifrac, factors = sampledf$region) 

###by season
set.seed(1911)
pairwise.adonis(type.wunifrac, factors = sampledf$season) 
```


Hill numbers - a framework to look at taxonomic, functional, and phylogenetic diversity

R package to do this: 
Li, (2018). hillR: taxonomic, functional, and phylogenetic diversity and similarity through Hill Numbers. Journal of Open Source Software, 3(31), 1041. https://doi.org/10.21105/joss.01041

```{r Hill Numbers - Full Dataset}
##note that with Hill #s, can account for species abundance w/ q =  (higher q = more sensitive to common species)
# q = 0; ignores species abundance, q = 1; weighs all species by abundance equally (Shannon's div), q = 2; common species get more weight (inverse Simpson)...

####Alpha diversity - Calculate taxonomic, functional, and phylogenetic diversity of each site####

# taxonomic alpha diversity

hill_taxa()

# functional alpha diversity

hill_func()

# phylogenetic alpha diversity

hill_phylo()

####Beta diversity - Calculate taxonomic, functional, and phylogenetic diversity across multiple sites####

# taxonomic beta diversity
hill_taxa_parti()

# functional beta diversity
hill_func_parti()

# phylogenetic beta diversity
hill_phylo_parti()

####Calculate pairwise taxonomic, functional, and phylogenetic diversity####

# pairwise taxonomic diversity
hill_taxa_parti_pairwise()

# pairwise functional diversity
hill_func_parti_pairwise()

# pairwise phylogenetic diversity
hill_phylo_parti_pairwise()
```
Following the full-dataset PERMANOVAs, we want to look more closely at our data, breaking the samples up by season/region/species to get a better idea of how the microbiome communities compare


```{r subsetting full gut dataset by season}

####subsetting data by season


#non-upwelling
ROHR_07_non_upwelling <- subset_samples(ROHR_07_obj_phylo_root, season == "non-upwelling")
ROHR_07_non_upwelling #155 samples (out of 312 samples total & 3218 taxa)

#upwelling
ROHR_07_upwelling <- subset_samples(ROHR_07_obj_phylo_root, season == "upwelling")
ROHR_07_upwelling #157 samples

#removings ASVs that are not present in any sample (if any)
ROHR_07_non_upwelling <- prune_taxa(taxa_sums(ROHR_07_non_upwelling) > 0, ROHR_07_non_upwelling)
ROHR_07_non_upwelling #2029 taxa and 155 samples

ROHR_07_upwelling <- prune_taxa(taxa_sums(ROHR_07_upwelling) > 0, ROHR_07_upwelling)
ROHR_07_upwelling #2131 taxa and 157 samples

##transforming to counts
ROHR_07_non_upwelling_prop = transform_sample_counts(ROHR_07_non_upwelling, function(x) x/sum(x))
ROHR_07_upwelling_prop = transform_sample_counts(ROHR_07_upwelling, function(x) x/sum(x))

# Save Phyloseq objects
saveRDS(ROHR_07_non_upwelling_prop, "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_non_upwelling_prop.rds")

saveRDS(ROHR_07_upwelling_prop, "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_upwelling_prop.rds")

# Read Phyloseq object
ROHR_07_non_upwelling_prop = readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_non_upwelling_prop.rds")
ROHR_07_upwelling_prop = readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_upwelling_prop.rds")

```

```{r PERMANOVA - samples broken down by season}
####running the PERMANOVA
metadata_seasonN <- as(sample_data(ROHR_07_non_upwelling), "data.frame")
metadata_seasonU <- as(sample_data(ROHR_07_upwelling), "data.frame")

####region/site + species

#non-upwelling
with(metadata_seasonN, adonis2(distance(ROHR_07_non_upwelling_prop, method="bray") ~ region/site + species,
        data = metadata_seasonN, strata = region))

#adonis2(formula = distance(ROHR_07_non_upwelling_prop, method = "bray") ~ region/site + species, data = metadata_seasonN, strata = region)
#             Df SumOfSqs      R2      F Pr(>F)    
#region        1    0.416 0.00598 1.3273  0.001 ***
#species       9   23.780 0.34136 8.4228  0.001 ***
#region:site   5    1.862 0.02672 1.1869  0.069 .  
#Residual    139   43.605 0.62594                  
#Total       154   69.663 1.00000 

#upwelling
with(metadata_seasonU, adonis2(distance(ROHR_07_upwelling_prop, method="bray") ~ region/site + species,
        data = metadata_seasonU, strata = region))

#adonis2(formula = distance(ROHR_07_upwelling_prop, method = "bray") ~ region/site + species, data = metadata_seasonU, strata = region)
#             Df SumOfSqs      R2       F Pr(>F)    
#region        1    0.696 0.00993  2.5509  0.001 ***
#species       9   29.176 0.41609 11.8816  0.001 ***
#region:site   7    2.322 0.03312  1.2159  0.035 *  
#Residual    139   37.925 0.54086                   
#Total       156   70.120 1.00000 
```

```{r subsetting full swab dataset by region}

####subsetting data by region

#Coiba (non-upwelling region)
ROHR_07_coiba <- subset_samples(ROHR_07_obj_phylo_root, region == "Coiba")
ROHR_07_coiba #158 samples

#Las Perlas (upwelling region)
ROHR_07_lasperlas <- subset_samples(ROHR_07_obj_phylo_root, region == "Las Perlas")
ROHR_07_lasperlas #154 samples

#removings ASVs that are not present in any sample (if any)
ROHR_07_coiba <- prune_taxa(taxa_sums(ROHR_07_coiba) > 0, ROHR_07_coiba)
ROHR_07_coiba #1624 taxa and 158 samples

ROHR_07_lasperlas <- prune_taxa(taxa_sums(ROHR_07_lasperlas) > 0, ROHR_07_lasperlas)
ROHR_07_lasperlas #2195 taxa and 154 samples

##transforming to counts
ROHR_07_coiba_prop = transform_sample_counts(ROHR_07_coiba, function(x) x/sum(x))
ROHR_07_lasperlas_prop = transform_sample_counts(ROHR_07_lasperlas, function(x) x/sum(x))

# Save Phyloseq objects
saveRDS(ROHR_07_coiba_prop, "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_coiba_prop.rds")

saveRDS(ROHR_07_lasperlas_prop, "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_lasperlas_prop.rds")

# Read Phyloseq object
ROHR_07_coiba_prop = readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_coiba_prop.rds")
ROHR_07_lasperlas_prop = readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_lasperlas_prop.rds")

```

```{r PERMANOVA - samples broken down by region}
####running the PERMANOVA
metadata_coiba <- as(sample_data(ROHR_07_coiba), "data.frame")
metadata_lasperlas <- as(sample_data(ROHR_07_lasperlas), "data.frame")

####season*site + species

#Coiba
with(metadata_coiba, adonis2(distance(ROHR_07_coiba_prop, method="bray") ~ season*site + species,
        data = metadata_coiba, strata = season))

#adonis2(formula = distance(ROHR_07_coiba_prop, method = "bray") ~ season * site + species, data = metadata_coiba, strata = season)
#             Df SumOfSqs      R2      F Pr(>F)    
#season        1    0.391 0.00550 1.3561  0.001 ***
#site          6    4.245 0.05962 2.4516  0.001 ***
#species       9   25.686 0.36075 9.8886  0.001 ***
#season:site   2    0.762 0.01070 1.3198  0.067 .  
#Residual    139   40.118 0.56343                  
#Total       157   71.203 1.00000 

#Las Perlas
with(metadata_lasperlas, adonis2(distance(ROHR_07_lasperlas_prop, method="bray") ~ season*site + species,
        data = metadata_lasperlas, strata = season))
#             Df SumOfSqs      R2      F Pr(>F)    
#season        1    1.309 0.01933 3.5861  0.001 ***
#site          2    1.198 0.01769 1.6412  0.001 ***
#species       9    9.347 0.13798 2.8447  0.001 ***
#season:site   1    0.397 0.00586 1.0865  0.280    
#Residual    152   55.491 0.81915                  
#Total       165   67.741 1.00000 

#adonis2(formula = distance(ROHR_07_lasperlas_prop, method = "bray") ~ season * site + species, data = metadata_lasperlas, strata = season)
#             Df SumOfSqs      R2      F Pr(>F)    
#season        1    0.530 0.00775 1.8009  0.001 ***
#site          3    2.379 0.03478 2.6950  0.001 ***
#species       9   24.193 0.35375 9.1364  0.001 ***
#season:site   1    0.392 0.00573 1.3322  0.133    
#Residual    139   40.896 0.59799                  
#Total       153   68.389 1.00000 

```

```{r subsetting season dataset by region}

####subsetting season data by region


#Coiba (non-upwelling)
ROHR_07_coiba_N <- subset_samples(ROHR_07_non_upwelling, region == "Coiba")
ROHR_07_coiba_N # 79 samples

#Coiba (upwelling)
ROHR_07_coiba_U <- subset_samples(ROHR_07_upwelling, region == "Coiba")
ROHR_07_coiba_U # 79 samples

#Las Perlas (non-upwelling)
ROHR_07_lasperlas_N <- subset_samples(ROHR_07_non_upwelling, region == "Las Perlas")
ROHR_07_lasperlas_N # 76 samples

#Las Perlas (upwelling)
ROHR_07_lasperlas_U <- subset_samples(ROHR_07_upwelling, region == "Las Perlas")
ROHR_07_lasperlas_U # 78 samples

#removings ASVs that are not present in any sample (if any)
ROHR_07_coiba_N <- prune_taxa(taxa_sums(ROHR_07_coiba_N) > 0, ROHR_07_coiba_N)
ROHR_07_coiba_N # 1256 taxa and 79 samples (out of full dataset --> 2131 taxa)

ROHR_07_coiba_U <- prune_taxa(taxa_sums(ROHR_07_coiba_U) > 0, ROHR_07_coiba_U)
ROHR_07_coiba_U # 1325 taxa and 79 samples

ROHR_07_lasperlas_N <- prune_taxa(taxa_sums(ROHR_07_lasperlas_N) > 0, ROHR_07_lasperlas_N)
ROHR_07_lasperlas_N #1721 taxa and 76 samples

ROHR_07_lasperlas_U <- prune_taxa(taxa_sums(ROHR_07_lasperlas_U) > 0, ROHR_07_lasperlas_U)
ROHR_07_lasperlas_U #1812 taxa and 78 samples

##transforming to counts
ROHR_07_coiba_N_prop = transform_sample_counts(ROHR_07_coiba_N, function(x) x/sum(x))
ROHR_07_coiba_U_prop = transform_sample_counts(ROHR_07_coiba_U, function(x) x/sum(x))
ROHR_07_lasperlas_N_prop = transform_sample_counts(ROHR_07_lasperlas_N, function(x) x/sum(x))
ROHR_07_lasperlas_U_prop = transform_sample_counts(ROHR_07_lasperlas_U, function(x) x/sum(x))

# Save Phyloseq objects
saveRDS(ROHR_07_coiba_N_prop, "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_coiba_N_prop.rds")

saveRDS(ROHR_07_lasperlas_N_prop, "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_lasperlas_N_prop.rds")

saveRDS(ROHR_07_coiba_U_prop, "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_coiba_U_prop.rds")

saveRDS(ROHR_07_lasperlas_U_prop, "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_lasperlas_U_prop.rds")

# Read Phyloseq object
ROHR_07_coiba_N_prop = readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_coiba_N_prop.rds")
ROHR_07_lasperlas_N_prop = readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_lasperlas_N_prop.rds")
ROHR_07_coiba_U_prop = readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_coiba_U_prop.rds")
ROHR_07_lasperlas_U_prop = readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_lasperlas_U_prop.rds")
```

```{r PERMANOVA - samples broken down by region & season}
####running the PERMANOVA
metadata_coiba_N <- as(sample_data(ROHR_07_coiba_N), "data.frame")
metadata_coiba_U <- as(sample_data(ROHR_07_coiba_U), "data.frame")
metadata_lasperlas_N <- as(sample_data(ROHR_07_lasperlas_N), "data.frame")
metadata_lasperlas_U <- as(sample_data(ROHR_07_lasperlas_U), "data.frame")

####species + site

#Coiba - non-upwelling
with(metadata_coiba_N, adonis2(distance(ROHR_07_coiba_N_prop, method="bray") ~ species + site,
        data = metadata_coiba_N))

#adonis2(formula = distance(ROHR_07_coiba_N_prop, method = "bray") ~ species + site, data = metadata_coiba_N)
#         Df SumOfSqs      R2      F Pr(>F)    
#species   9   14.099 0.39644 5.0445  0.001 ***
#site      2    0.658 0.01851 1.0598  0.330    
#Residual 67   20.806 0.58505                  
#Total    78   35.563 1.00000 

#Coiba - upwelling
with(metadata_coiba_U, adonis2(distance(ROHR_07_coiba_U_prop, method="bray") ~ species + site,
        data = metadata_coiba_U))

#adonis2(formula = distance(ROHR_07_coiba_U_prop, method = "bray") ~ species + site, data = metadata_coiba_U)
#         Df SumOfSqs      R2      F Pr(>F)    
#species   9   16.937 0.48051 7.2411  0.001 ***
#site      6    1.938 0.05497 1.2427  0.035 *  
#Residual 63   16.373 0.46451                  
#Total    78   35.249 1.00000 

#Las Perlas - non-upwelling
with(metadata_lasperlas_N, adonis2(distance(ROHR_07_lasperlas_N_prop, method="bray") ~ species + site,
        data = metadata_lasperlas_N))

#adonis2(formula = distance(ROHR_07_lasperlas_N_prop, method = "bray") ~ species + site, data = metadata_lasperlas_N)
#         Df SumOfSqs      R2      F Pr(>F)    
#species   9   13.085 0.38847 4.6891  0.001 ***
#site      3    1.065 0.03161 1.1445  0.163    
#Residual 63   19.534 0.57992                  
#Total    75   33.684 1.00000  

#Las Perlas - upwelling
with(metadata_lasperlas_U, adonis2(distance(ROHR_07_lasperlas_U_prop, method="bray") ~ species + site,
        data = metadata_lasperlas_U))

#adonis2(formula = distance(ROHR_07_lasperlas_U_prop, method = "bray") ~ species + site, data = metadata_lasperlas_U)
#         Df SumOfSqs      R2     F Pr(>F)    
#species   9   16.316 0.47741 6.893  0.001 ***
#site      1    0.239 0.00699 0.908  0.556    
#Residual 67   17.621 0.51560                 
#Total    77   34.175 1.00000  

```


Using unconstrained ordinations to visualize our data, following:
https://deneflab.github.io/MicrobeMiseq/demos/mothur_2_phyloseq.html
```{r unconstrained ordinations - full dataset}
theme_set(theme_bw())

# Ordinate
ROHR_07_pcoa <- ordinate(
  physeq = ROHR_07_obj_clean_prop, 
  method = "PCoA", 
  distance = "bray"
)

# Plot - color = fish host species, shape = region (coiba or las perlas)
plot_ordination(
  physeq = ROHR_07_obj_clean_prop,
  ordination = ROHR_07_pcoa,
  color = "species",
  shape = "season",
  title = "PCoA of Tropical Eastern Pacific Fish Gut Microbiome Communities"
) + 
  scale_color_brewer(palette = "Paired" ) +
  geom_point(aes(color = species), alpha = 0.7, size = 3) +
  geom_point(colour = "grey90", size = 1)

# Same plot as above, except faceted by season & region
#Plot - color = fish host species
plot_ordination(
  physeq = ROHR_07_obj_clean_prop,
  ordination = ROHR_07_pcoa,
  color = "species",
  title = "PCoA of Tropical Eastern Pacific Fish Gut Microbiome Communities"
) + 
  scale_color_brewer(palette = "Paired" ) +
  geom_point(aes(color = species), alpha = 0.7, size = 3) +
  geom_point(colour = "grey90", size = 1) +
  facet_wrap(~region_season)

# Plot with both species & taxa - color = fish host species, shape = region (coiba or las perlas)
plot_ordination(
  physeq = ROHR_07_obj_clean_prop,
  ordination = ROHR_07_pcoa,
  color = "species",
  shape = "region",
  type = "split",
  title = "PCoA of Tropical Eastern Pacific Fish Gut Microbiome Communities"
) + 
  scale_color_brewer(palette = "Paired" ) +
  geom_point(aes(color = species), alpha = 0.7, size = 3) +
  geom_point(colour = "grey90", size = 1)


  
```

```{r unconstrained ordinations - Coiba}

# Ordinate
ROHR_07_pcoa_coiba <- ordinate(
  physeq = ROHR_07_coiba_prop, 
  method = "PCoA", 
  distance = "bray"
)

# Plot - color = fish host species, shape = season
plot_ordination(
  physeq = ROHR_07_coiba_prop,
  ordination = ROHR_07_pcoa_coiba,
  color = "species",
  shape = "season",
  title = "PCoA of Coiba Fish Gut Microbiome Communities"
) + 
  scale_color_brewer(palette = "Paired" ) +
  geom_point(aes(color = species), alpha = 0.7, size = 4) +
  geom_point(colour = "grey90", size = 1.5)

# Plot - color = fish host species, shape = site
plot_ordination(
  physeq = ROHR_07_coiba_prop,
  ordination = ROHR_07_pcoa_coiba,
  color = "species",
  shape = "site",
  title = "PCoA of Coiba Fish Gut Microbiome Communities"
) + 
  scale_color_brewer(palette = "Paired" ) +
  geom_point(aes(color = species), alpha = 0.7, size = 4) +
  geom_point(colour = "grey90", size = 1.5)
```


```{r unconstrained ordinations - Las Perlas}
# Ordinate
ROHR_07_pcoa_lasperlas <- ordinate(
  physeq = ROHR_07_lasperlas_prop, 
  method = "PCoA", 
  distance = "bray"
)

# Plot - color = fish host species, shape = season
plot_ordination(
  physeq = ROHR_07_lasperlas_prop,
  ordination = ROHR_07_pcoa_lasperlas,
  color = "species",
  shape = "season",
  title = "PCoA of Las Perlas Fish Gut Microbiome Communities"
) + 
  scale_color_brewer(palette = "Paired" ) +
  geom_point(aes(color = species), alpha = 0.7, size = 4) +
  geom_point(colour = "grey90", size = 1.5)

# Plot - color = fish host species, shape = site
plot_ordination(
  physeq = ROHR_07_lasperlas_prop,
  ordination = ROHR_07_pcoa_lasperlas,
  color = "species",
  shape = "site",
  title = "PCoA of Las Perlas Fish Gut Microbiome Communities"
) + 
  scale_color_brewer(palette = "Paired" ) +
  geom_point(aes(color = species), alpha = 0.7, size = 4) +
  geom_point(colour = "grey90", size = 1.5)
```



```{r unconstrained ordinations - Coiba - by season}

# Ordinate - Non-upwelling
ROHR_07_pcoa_coiba_N <- ordinate(
  physeq = ROHR_07_coiba_N_prop, 
  method = "PCoA", 
  distance = "bray"
)

# Plot - color = fish host species, shape = site
plot_ordination(
  physeq = ROHR_07_coiba_N_prop,
  ordination = ROHR_07_pcoa_coiba_N,
  color = "species",
  shape = "site",
  title = "PCoA of Coiba Fish Gut Microbiome Communities During Non-Upwelling Season"
) + 
  scale_color_brewer(palette = "Paired" ) +
  geom_point(aes(color = species), alpha = 0.7, size = 4) +
  geom_point(colour = "grey90", size = 1.5)


# Ordinate - Upwelling
ROHR_07_pcoa_coiba_U <- ordinate(
  physeq = ROHR_07_coiba_U_prop, 
  method = "PCoA", 
  distance = "bray"
)

# Plot - color = fish host species, shape = site
plot_ordination(
  physeq = ROHR_07_coiba_U_prop,
  ordination = ROHR_07_pcoa_coiba_U,
  color = "species",
  shape = "site",
  title = "PCoA of Coiba Fish Gut Microbiome Communities During Upwelling Season"
) + 
  scale_color_brewer(palette = "Paired" ) +
  geom_point(aes(color = species), alpha = 0.7, size = 4) +
  geom_point(colour = "grey90", size = 1.5)
```
```{r unconstrained ordinations - Las Perlas - by season}
# Ordinate - non-upwelling
ROHR_07_pcoa_lasperlas_N <- ordinate(
  physeq = ROHR_07_lasperlas_N_prop, 
  method = "PCoA", 
  distance = "bray"
)

# Plot - color = fish host species, shape = site
plot_ordination(
  physeq = ROHR_07_lasperlas_N_prop,
  ordination = ROHR_07_pcoa_lasperlas_N,
  color = "species",
  shape = "site",
  title = "PCoA of Las Perlas Fish Gut Microbiome Communities During Non-Upwelling Season"
) + 
  scale_color_brewer(palette = "Paired" ) +
  geom_point(aes(color = species), alpha = 0.7, size = 4) +
  geom_point(colour = "grey90", size = 1.5)

# Ordinate - upwelling
ROHR_07_pcoa_lasperlas_U <- ordinate(
  physeq = ROHR_07_lasperlas_U_prop, 
  method = "PCoA", 
  distance = "bray"
)

# Plot - color = fish host species, shape = site
plot_ordination(
  physeq = ROHR_07_lasperlas_U_prop,
  ordination = ROHR_07_pcoa_lasperlas_U,
  color = "species",
  shape = "site",
  title = "PCoA of Las Perlas Fish Gut Microbiome Communities During Upwelling Season"
) + 
  scale_color_brewer(palette = "Paired" ) +
  geom_point(aes(color = species), alpha = 0.7, size = 4) +
  geom_point(colour = "grey90", size = 1.5)


```



NMDS (Non-metric multidimensional scaling)
- another ordination technique
- calculates limited # of axes (iterative approach - tries to find best solution)
- important to look at stress = goodness of fit of the regression - i.e., how close the plotted points in ordination space are to the rank order of distances in the original distance matrix

more NMDS info: http://strata.uga.edu/8370/lecturenotes/multidimensionalScaling.html

```{r NMDS}
set.seed(1)

# Ordinate
ROHR_07_nmds <- ordinate(
  physeq = ROHR_07_obj_clean_prop, 
  method = "NMDS", 
  distance = "bray"
)

plot_ordination(
  physeq = ROHR_07_obj_clean_prop,
  ordination = ROHR_07_nmds,
  color = "species",
  shape = "region",
  title = "NMDS of Tropical Eastern Pacific Fish Gut Microbial Communities") + 
  scale_color_brewer(palette = "Paired") +
  geom_point(aes(color = species), alpha = 0.7, size = 3) +
  geom_point(colour = "grey90", size = 1) 

##stress ~0.27 (relatively high, from what I understand - but makes sense since we are working with the full dataset)
```

```{r NMDS - by region - Coiba}
set.seed(1)

# Ordinate
ROHR_07_nmds_coiba <- ordinate(
  physeq = ROHR_07_coiba_prop, 
  method = "NMDS", 
  distance = "bray"
)

plot_ordination(
  physeq = ROHR_07_coiba_prop,
  ordination = ROHR_07_nmds_coiba,
  color = "species",
  shape = "season",
  title = "NMDS of Coiba Fish Gut Microbial Communities") + 
  scale_color_brewer(palette = "Paired") +
  geom_point(aes(color = species), alpha = 0.7, size = 3) +
  geom_point(colour = "grey90", size = 1) 

#still seeing similarly high (0.26) levels of stress
```
```{r NMDS - by region - Las Perlas}
set.seed(1)

# Ordinate
ROHR_07_nmds_lasperlas <- ordinate(
  physeq = ROHR_07_lasperlas_prop, 
  method = "NMDS", 
  distance = "bray"
)

plot_ordination(
  physeq = ROHR_07_lasperlas_prop,
  ordination = ROHR_07_nmds_lasperlas,
  color = "species",
  shape = "season",
  title = "NMDS of Las Perlas Fish Gut Microbial Communities") + 
  scale_color_brewer(palette = "Paired") +
  geom_point(aes(color = species), alpha = 0.7, size = 3) +
  geom_point(colour = "grey90", size = 1) 

#still seeing similarly high (0.27) levels of stress
```

Now, do constrained ordinations to look at how environmental variables are associated with these changes in community composition
(following https://deneflab.github.io/MicrobeMiseq/demos/mothur_2_phyloseq.html)

```{r constrained ordination - full dataset}
 ###fix this so that the arrows are more legible (remove arrows for species & multiple sites w/in region (perhaps would help to merge dataset by region first?))   

ROHR_07_bray <- phyloseq::distance(physeq = ROHR_07_obj_clean_prop, method = "bray")

                            
# CAP ordinate
cap_ord <- ordinate(
    physeq = ROHR_07_obj_clean_prop, 
    method = "CAP",
    distance = ROHR_07_bray,
    formula = ~ species*region:site*season
)

# CAP plot
cap_plot <- plot_ordination(
  physeq = ROHR_07_obj_clean_prop, 
  ordination = cap_ord, 
    color = "species", 
    axes = c(1,2)
) + 
    aes(shape = region) + 
    geom_point(aes(colour = species), alpha = 0.4, size = 4) + 
    geom_point(colour = "grey90", size = 1.5) + 
    scale_color_brewer(palette = "Paired"
    )


# Now add the environmental variables as arrows
arrowmat <- vegan::scores(cap_ord, display = "bp")

# Add labels, make a data.frame
arrowdf <- data.frame(labels = rownames(arrowmat), arrowmat)

# Define the arrow aesthetic mapping
arrow_map <- aes(xend = CAP1, 
    yend = CAP2, 
    x = 0, 
    y = 0, 
    shape = NULL, 
    color = NULL, 
    label = labels)

label_map <- aes(x = 1.3 * CAP1, 
    y = 1.3 * CAP2, 
    shape = NULL, 
    color = NULL, 
    label = labels)

arrowhead = arrow(length = unit(0.02, "npc"))

# Make a new graphic
cap_plot + 
  geom_segment(
    mapping = arrow_map, 
    size = .5, 
    data = arrowdf, 
    color = "gray", 
    arrow = arrowhead
  ) + 
  geom_text(
    mapping = label_map, 
    size = 4,  
    data = arrowdf, 
    show.legend = FALSE
  )

#anova of constrained axes used in ordination
anova(cap_ord)
```


Following https://deneflab.github.io/MicrobeMiseq/demos/mothur_2_phyloseq.html
```{r prep data visualize most abundant taxa}

# melt to long format (for ggploting) 
# prune out phyla below 2% in each sample

ROHR_07_phylum <- ROHR_07_obj_phylo_root %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  filter(Abundance > 0.02) %>%                         # Filter out low abundance taxa (remove those that contribute less                                                          than 2% of relative abundance of each sample)
  arrange(Phylum)                                      # Sort data frame alphabetically by phylum

```


```{r stacked barplot to visualize most abundant taxa - broken into regions}
#pick colors for bars
c25 <- c(
  "dodgerblue2", "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "black", "gold1",
  "skyblue2", "#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", # lt purple
  "#FDBF6F", # lt orange
  "gray70", "khaki2",
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown")

###use colors assigned to ROHR_06 (skin) to match --> but add color for those unique to gut
c3 <- c("yellow3","darkorange4", "brown")

unique(ROHR_06_phylum$Phylum) #skin (from ROHR_06_Swab_stats.Rmd --> need to run this code first --> will add code chunk to this doc later so they can be run independently)
unique(ROHR_07_phylum$Phylum) #gut
#unique to gut
#"WPS-2", "Spirochaetota", "Deferribacterota"

names(c25) = c(unique(ROHR_06_phylum$Phylum), "WPS-2", "Spirochaetota", "Deferribacterota")
c25

# Plot 
##can add --> position = "fill" to geom_bar() to make bars take up full height

ggplot(ROHR_07_phylum, aes(x = species, y = Abundance, fill = fct_reorder(Phylum, Abundance))) + 
  facet_grid(region~.) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c25) +
  scale_x_discrete(
    breaks = c("Abudefduf concolor", "Prionurus laticlavius", "Chaetodon humeralis", "Acanthurus xanthopterus", "Cephalopholis panamensis", "Johnrandallia nigrirostris", "Abudefduf troschelii", "Microspathodon dorsalis", "Epinephelus labriformis", "Cephalopholis colonus"),
    labels = c("A. concolor", "P. laticlavius", "C. humeralis", "A. xanthopterus", "C. panamensis", "J. nigrirostris", "A. troschelii", "M. dorsalis", "E. labriformis", "C. colonus"), 
    drop = FALSE
  ) +
  #
  guides(fill = guide_legend(reverse = TRUE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Phyla > 2%) \n") +
  ggtitle("Phylum Composition of TEP Fish Gut Microbial Communities by Region") 


#stacked barplot to visualize most abundant taxa - broken into seasons
##can add --> position = "fill" to geom_bar() to make bars take up full height

ggplot(ROHR_07_phylum, aes(x = species, y = Abundance, fill = fct_reorder(Phylum, Abundance))) + 
  facet_grid(season~.) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c25) +
  scale_x_discrete(
    breaks = c("Abudefduf concolor", "Prionurus laticlavius", "Chaetodon humeralis", "Acanthurus xanthopterus", "Cephalopholis panamensis", "Johnrandallia nigrirostris", "Abudefduf troschelii", "Microspathodon dorsalis", "Epinephelus labriformis", "Cephalopholis colonus"),
    labels = c("A. concolor", "P. laticlavius", "C. humeralis", "A. xanthopterus", "C. panamensis", "J. nigrirostris", "A. troschelii", "M. dorsalis", "E. labriformis", "C. colonus"), 
    drop = FALSE
  ) +
  #
  guides(fill = guide_legend(reverse = TRUE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Phyla > 2%) \n") +
  ggtitle("Phylum Composition of TEP Fish Gut Microbial Communities by Season") 

#stacked barplot to visualize most abundant taxa - broken into seasons & regions
##can add --> position = "fill" to geom_bar() to make bars take up full height

ggplot(ROHR_07_phylum, aes(x = species, y = Abundance, fill = fct_reorder(Phylum, Abundance))) + 
  facet_grid(region_season~.) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c25) +
  scale_x_discrete(
    breaks = c("Abudefduf concolor", "Prionurus laticlavius", "Chaetodon humeralis", "Acanthurus xanthopterus", "Cephalopholis panamensis", "Johnrandallia nigrirostris", "Abudefduf troschelii", "Microspathodon dorsalis", "Epinephelus labriformis", "Cephalopholis colonus"),
    labels = c("A. concolor", "P. laticlavius", "C. humeralis", "A. xanthopterus", "C. panamensis", "J. nigrirostris", "A. troschelii", "M. dorsalis", "E. labriformis", "C. colonus"), 
    drop = FALSE
  ) +
  #
  guides(fill = guide_legend(reverse = TRUE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Phyla > 2%) \n") +
  ggtitle("Phylum Composition of TEP Fish Gut Microbial Communities by Season & Region ")

##glammed up version of plot above
#same plot w/o title

ggplot(ROHR_07_phylum, aes(x = species, y = Abundance, fill = fct_reorder(Phylum, Abundance))) + 
  facet_grid(region_season~.) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c25, name="Microbial Phyla")+
  scale_x_discrete(
    breaks = c("Abudefduf concolor", "Prionurus laticlavius", "Chaetodon humeralis", "Acanthurus xanthopterus", "Cephalopholis panamensis", "Johnrandallia nigrirostris", "Abudefduf troschelii", "Microspathodon dorsalis", "Epinephelus labriformis", "Cephalopholis colonus"),
    labels = c("A. concolor", "P. laticlavius", "C. humeralis", "A. xanthopterus", "C. panamensis", "J. nigrirostris", "A. troschelii", "M. dorsalis", "E. labriformis", "C. colonus"), 
    drop = FALSE
  ) +
  #
  guides(fill = guide_legend(reverse = TRUE, keywidth = 1, keyheight = 1, ncol = 1)) +
  ylab("Relative Abundance (Phyla > 2%) \n") +
  xlab("Host Species") +
  scale_y_continuous(labels = scales::percent)+
  theme(axis.text.x = element_text(angle = 15, hjust = 1, size = 14, face = "italic"), #change size & orientation of species names (x-axis) & fix overlap
        axis.title.x = element_text(size = 16), #change size of x-axis label (Species)
        axis.text.y = element_text(size = 10), #change size of y-axis #s
        axis.title.y = element_text(size = 16), #change size of y-axis title
        legend.title = element_text(size = 16), #change size of legend title (microbial phyla)
        legend.text = element_text(size = 11), #change size of legend text (microbial taxa)
        strip.text.y = element_text(size = 13)) #change size of each box's label (e.g., Coiba upwelling)

# stacked barplot to visualize most abundant taxa - broken into site

ggplot(ROHR_07_phylum, aes(x = species, y = Abundance, fill = fct_reorder(Phylum, Abundance))) + 
  facet_grid(site~.) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c25) +
  scale_x_discrete(
    breaks = c("Abudefduf concolor", "Prionurus laticlavius", "Chaetodon humeralis", "Acanthurus xanthopterus", "Cephalopholis panamensis", "Johnrandallia nigrirostris", "Abudefduf troschelii", "Microspathodon dorsalis", "Epinephelus labriformis", "Cephalopholis colonus"),
    labels = c("A. concolor", "P. laticlavius", "C. humeralis", "A. xanthopterus", "C. panamensis", "J. nigrirostris", "A. troschelii", "M. dorsalis", "E. labriformis", "C. colonus"), 
    drop = FALSE
  ) +
  #
  guides(fill = guide_legend(reverse = TRUE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Phyla > 2%) \n") +
  ggtitle("Phylum Composition of TEP Fish Gut Microbial Communities by Sampling Site") 

```

From these abudance graphs, it is quite clear that each fish hosts a significantly different gut microbiome, with certain species such as P. laticlavius having a particularly rich and relatively even microbiome while most are dominated by bacteria from a few main phyla, namely Proteobacteria (20-80% relative abundance), Fusobacteriota, Firmicutes, and Bacteroidota (only highly represented in P. laticlavius & J. nigrirostris). Compared to the skin samples, there is less taxonomic diversity: while the most abundant (>2% rel abund) gut microbes of these fish come from 16 phyla, the skin of these same fish had 22 phyla represented. In both skin & gut, Proteobacteria dominated, but contrary to the gut microbiome, the other top phyla for skin were Bacteroidota (across all fish), Cyanobacteria, and Actinobacteriota. Additionally, interspecific differences appear to be larger for the gut microbiome than the skin microbiome. There seem to be some region-specific differences (e.g., presence of deinococcota in Coiba M. dorsalis)



Taking a quick look at the Caribbean fish skin microbiome
```{r Caribbean skin microbiome preliminary visualizations}
#read RDS file

ROHR_07_skin_phylo_root <- readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_skin_with_phylogenetic_tree_rooted.rds")

```

```{r prep data visualize most abundant taxa - Caribbean skin}

# melt to long format (for ggploting) 
# prune out phyla below 2% in each sample

ROHR_07_phylum_skin <- ROHR_07_skin_phylo_root %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  filter(Abundance > 0.02) %>%                         # Filter out low abundance taxa (remove those that contribute less                                                          than 2% of relative abundance of each sample)
  arrange(Phylum)                                      # Sort data frame alphabetically by phylum

```


```{r Caribbean skin stacked barplot to visualize most abundant taxa}

unique(ROHR_06_phylum$Phylum) #skin (from ROHR_06_Swab_stats.Rmd --> need to run this code first --> will add code chunk to this doc later so they can be run independently)
unique(ROHR_07_phylum_skin$Phylum) #skin Car
#unique to Caribbean skin = none

#rename SAR324 clade, as done for TEP samples
ROHR_07_phylum_skin["Phylum"][ROHR_07_phylum_skin["Phylum"] == "SAR324 clade(Marine group B)"] <- "SAR324 clade"

#check this worked
unique(ROHR_07_phylum_skin$Phylum)

#look at species names for Caribbean
unique(ROHR_07_phylum_skin$species)

#"Anisotremus virginicus", "Abudefduf taurus", "Abudefduf saxatilis", "Microspathodon chrysurus", "Chaetodon ocellatus", "Chaetodon capistratus", "Chaetodon striatus" 

#reorder samples so that fish are grouped by diet
ROHR_07_phylum_skin <- ROHR_07_phylum_skin %>%
  mutate(species = fct_relevel(species, "Abudefduf taurus", "Microspathodon chrysurus", "Chaetodon ocellatus", "Chaetodon capistratus", "Chaetodon striatus", "Anisotremus virginicus", "Abudefduf saxatilis"))

##glammed up version of plot
#w/o title

ggplot(ROHR_07_phylum_skin, aes(x = species, y = Abundance, fill = fct_reorder(Phylum, Abundance))) + 
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c25, name="Microbial Phyla")+
  scale_x_discrete(
    breaks = c("Abudefduf taurus", "Microspathodon chrysurus", "Chaetodon ocellatus", "Chaetodon capistratus", "Chaetodon striatus", "Anisotremus virginicus", "Abudefduf saxatilis"),
    labels = c("A. taurus", "M. chrysurus", "C. ocellatus", "C. capistratus", "C. striatus", "A. virginicus", "A. saxatilis"), 
    drop = FALSE
  ) +
  #
  guides(fill = guide_legend(reverse = TRUE, keywidth = 1, keyheight = 1, ncol = 1)) +
  ylab("Relative Abundance (Phyla > 2%) \n") +
  xlab("Host Species") +
  scale_y_continuous(labels = scales::percent)+
  theme(axis.text.x = element_text(angle = 15, hjust = 1, size = 14, face = "italic"), #change size & orientation of species names (x-axis) & fix overlap
        axis.title.x = element_text(size = 16), #change size of x-axis label (Species)
        axis.text.y = element_text(size = 10), #change size of y-axis #s
        axis.title.y = element_text(size = 16), #change size of y-axis title
        legend.title = element_text(size = 16), #change size of legend title (microbial phyla)
        legend.text = element_text(size = 11)) #change size of legend text (microbial taxa)
```

```{prelim sister species pairs comparison}

##read in new metadata for ROHR07 to match ROHR06 (remove two columns)
#creating path to the .csv 
ROHR_07_sam <- "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_7_metadata_new.csv"
#reading the sample data sheet
ROHR_07_metadata <- read.csv(ROHR_07_sam, header = TRUE, sep = ",", row.names = 1)

CAR_skin_samples <- sample_data(ROHR_07_metadata)

CAR_skin_tax <- tax_table(ROHR_07_skin_phylo_root)

CAR_skin_OTU <- otu_table(ROHR_07_skin_phylo_root)

#new phyloseq object
CARsis <- phyloseq(CAR_skin_samples, CAR_skin_OTU, CAR_skin_tax)

#remove non-sister species (no porkfish in TEP dataset for now)
ROHR_07_Sisters <- prune_samples(sample_data(CARsis)$species != "Anisotremus virginicus", CARsis)

#creating path to the .csv 
ROHR_06_sam <- "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_06_trimmed/ROHR_6_metadata.csv"
#reading the sample data sheet
ROHR_06_metadata <- read.csv(ROHR_06_sam, header = TRUE, sep = ",", row.names = 1)


TEP_skin_samples <- sample_data(ROHR_06_metadata)

TEP_skin_tax <- tax_table(ROHR_06_obj_phylo_root)

TEP_skin_OTU <- otu_table(ROHR_06_obj_phylo_root)

TEPsis <- phyloseq(TEP_skin_samples, TEP_skin_OTU, TEP_skin_tax)


ROHR_06_Sisters <- prune_samples(sample_data(TEPsis)$species == c("Abudefduf concolor", "Abudefduf troschelii", "Chaetodon humeralis", "Microspathodon dorsalis"), TEPsis)

#merge TEP & CAr datasets
merged_skin_TEP_CAR = merge_phyloseq(ROHR_07_Sisters, ROHR_06_Sisters)

#save merged data
saveRDS(merged_skin_TEP_CAR, "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/sisters_merged_TEP_CAR_prelim.rds")

#save subsetted data
saveRDS(ROHR_06_Sisters, "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_06_trimmed/ROHR_06_sister_only.rds")
saveRDS(ROHR_07_Sisters, "~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_sisters_only.rds")

#read subsetted data
ROHR_06_Sisters <- readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_06_trimmed/ROHR_06_sisters_only.rds")

ROHR_07_Sisters <- readRDS("~/OneDrive - McGill University/McGill/Lab_Docs/Fish_Data/Fish_2022/2021-2022_TEP_Fish/16S_Summer_2022/ROHR_07_trimmed/ROHR_07_Sisters_only.rds")

```



```{r prep data visualize most abundant taxa - sister pairs}

# melt to long format (for ggploting) 
# prune out phyla below 2% in each sample

ROHR_sisters_phylum_skin <- merged_skin_TEP_CAR %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  filter(Abundance > 0.02) %>%                         # Filter out low abundance taxa (remove those that contribute less                                                          than 2% of relative abundance of each sample)
  arrange(Phylum)                                      # Sort data frame alphabetically by phylum

```


```{r Caribbean skin stacked barplot to visualize most abundant taxa}

unique(ROHR_sisters_phylum_skin$Phylum)


#rename SAR324 clade & marinimicrobia
ROHR_sisters_phylum_skin["Phylum"][ROHR_sisters_phylum_skin["Phylum"] == "SAR324 clade(Marine group B)"] <- "SAR324 clade"

ROHR_sisters_phylum_skin["Phylum"][ROHR_sisters_phylum_skin["Phylum"] == "Marinimicrobia (SAR406 clade)"] <- "Marinimicrobia"

#check this worked
unique(ROHR_sisters_phylum_skin$Phylum)

#look at species names
unique(ROHR_sisters_phylum_skin$species)

# "Abudefduf concolor", "Microspathodon dorsalis", "Abudefduf troschelii", "Chaetodon humeralis", "Abudefduf taurus", "Abudefduf saxatilis", "Microspathodon chrysurus", "Chaetodon ocellatus", "Chaetodon capistratus", "Chaetodon striatus" 

#reorder samples so that fish are grouped by diet & sister pairs
ROHR_sisters_phylum_skin <- ROHR_sisters_phylum_skin %>%
  mutate(species = fct_relevel(species, "Abudefduf concolor", "Abudefduf taurus", "Microspathodon dorsalis", "Microspathodon chrysurus", "Chaetodon humeralis", "Chaetodon ocellatus", "Chaetodon capistratus", "Chaetodon striatus", "Abudefduf troschelii", "Abudefduf saxatilis"))

#pick colors for bars
c25 <- c(
  "dodgerblue2", "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "black", "gold1",
  "skyblue2", "#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", # lt purple
  "#FDBF6F", # lt orange
  "gray70", "khaki2",
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown")

##assign names to this vector of hex color values: names have to correspond to the labels of the groups that we want to colorize

names(c25) = unique(ROHR_sisters_phylum_skin$Phylum)
print(c25)

#Plot w/o title

ggplot(ROHR_sisters_phylum_skin, aes(x = species, y = Abundance, fill = fct_reorder(Phylum, Abundance))) + 
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c25, name="Microbial Phyla")+
  scale_x_discrete(
    breaks = c("Abudefduf concolor", "Abudefduf taurus", "Microspathodon dorsalis", "Microspathodon chrysurus", "Chaetodon humeralis", "Chaetodon ocellatus", "Chaetodon capistratus", "Chaetodon striatus", "Abudefduf troschelii", "Abudefduf saxatilis"),
    labels = c("A. concolor", "A. taurus", "M. dorsalis","M. chrysurus", "C. humeralis","C. ocellatus", "C. capistratus", "C. striatus", "A. troschelii", "A. saxatilis"), 
    drop = FALSE
  ) +
  #
  guides(fill = guide_legend(reverse = TRUE, keywidth = 1, keyheight = 1, ncol = 1)) +
  ylab("Relative Abundance (Phyla > 2%) \n") +
  xlab("Host Species") +
  scale_y_continuous(labels = scales::percent)+
  theme(axis.text.x = element_text(angle = 15, hjust = 1, size = 14, face = "italic"), #change size & orientation of species names (x-axis) & fix overlap
        axis.title.x = element_text(size = 16), #change size of x-axis label (Species)
        axis.text.y = element_text(size = 10), #change size of y-axis #s
        axis.title.y = element_text(size = 16), #change size of y-axis title
        legend.title = element_text(size = 16), #change size of legend title (microbial phyla)
        legend.text = element_text(size = 11)) #change size of legend text (microbial taxa)

#Plot w/o title --> facet by ocean

ggplot(ROHR_sisters_phylum_skin, aes(x = species, y = Abundance, fill = fct_reorder(Phylum, Abundance))) + 
  geom_bar(stat = "identity", position = "fill") +
  facet_grid(ocean~.) +
  scale_fill_manual(values = c25, name="Microbial Phyla")+
  scale_x_discrete(
    breaks = c("Abudefduf concolor", "Abudefduf taurus", "Microspathodon dorsalis", "Microspathodon chrysurus", "Chaetodon humeralis", "Chaetodon ocellatus", "Chaetodon capistratus", "Chaetodon striatus", "Abudefduf troschelii", "Abudefduf saxatilis"),
    labels = c("A. concolor", "A. taurus", "M. dorsalis","M. chrysurus", "C. humeralis","C. ocellatus", "C. capistratus", "C. striatus", "A. troschelii", "A. saxatilis"), 
    drop = FALSE
  ) +
  #
  guides(fill = guide_legend(reverse = TRUE, keywidth = 1, keyheight = 1, ncol = 1)) +
  ylab("Relative Abundance (Phyla > 2%) \n") +
  xlab("Host Species") +
  scale_y_continuous(labels = scales::percent)+
  theme(axis.text.x = element_text(angle = 15, hjust = 1, size = 14, face = "italic"), #change size & orientation of species names (x-axis) & fix overlap
        axis.title.x = element_text(size = 16), #change size of x-axis label (Species)
        axis.text.y = element_text(size = 10), #change size of y-axis #s
        axis.title.y = element_text(size = 16), #change size of y-axis title
        legend.title = element_text(size = 16), #change size of legend title (microbial phyla)
        legend.text = element_text(size = 11)) #change size of legend text (microbial taxa)


####will want to re-run this having "rarefied" or otherwise standardized the # of samples across oceans (or limited to just Gulf of Panama) as there are many more replicates (>2X) for TEP than Caribbean
```